---
title: 'Composites'
description: >
  Composite variables with lavaan's `<~` operator: what they are, how they
  differ from defined parameters, and how INLAvaan fits them.
vignette: >
  %\VignetteIndexEntry{Composites}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
bibliography: ../refs.bib
toc: true
---

```{r}
#| label: setup
#| include: false
library(INLAvaan)
library(lavaan)
set.seed(2026)
```

## What is a composite?

A composite is a weighted sum of observed variables. In lavaan syntax,
`C <~ x1 + x2 + x3` defines
$$
C_i = w_1 x_{i1} + w_2 x_{i2} + w_3 x_{i3} = \mathbf w^\top \mathbf x_i ,
$$
with the weights estimated and the first fixed at 1 to set the scale. Unlike a
common factor, a composite has no disturbance term. That is, it summarises its
indicators instead of explaining them.

A composite is not a defined parameter either. A `:=` parameter such as
`ind := a*b` is a function of the parameters---one value per posterior draw,
with no effect on the likelihood. A composite has a value for every person,
and it changes the model-implied covariance matrix
$\boldsymbol\Sigma(\boldsymbol\theta)$ through its weights.

## One outcome or several

With a single outcome, the composite model is a regression in disguise:
`C <~ x1 + x2 + x3; x4 ~ C` fits exactly as `x4 ~ x1 + x2 + x3` does, with
slope $\gamma_1$ on $C$ and weights $w_j = \gamma_j / \gamma_1$. Further
outcomes see the indicators only through $C$, so their regression coefficients
on the indicators must be proportional. Each extra outcome adds $p - 1$
degrees of freedom for $p$ indicators, and the model below has two.

## Not PLS-SEM

Partial least squares path modelling (PLS-SEM) is Wold's iterative estimation
algorithm, which has no likelihood. lavaan instead fits the composite model of
@dijkstra2017perfect by maximum likelihood, the same model that the
Henseler--Ogasawara specification [@schuberth2023henselerogasawara] reaches
through latent variables. INLAvaan puts priors on this likelihood, so the
result is Bayesian composite SEM, not Bayesian PLS.

## An example

We regress two verbal tests of the Holzinger and Swineford data (`x4` and
`x5`) on a composite of the three visual tests (`x1` to `x3`).

```{r}
model <- "
  C <~ x1 + x2 + x3
  x4 ~ C
  x5 ~ C
  x4 ~~ x5
"
utils::data("HolzingerSwineford1939", package = "lavaan")

fit <- asem(model, HolzingerSwineford1939)
summary(fit)
```

The marker `x1` is fixed at 1. The variance of `C` is not a parameter but the
sample variance of $\mathbf w^\top \mathbf x$ at the posterior means, so it has
no posterior SD. The posterior means sit close to lavaan's maximum likelihood
estimates.

```{r}
fit_ml <- sem(model, HolzingerSwineford1939)
round(cbind(INLAvaan = coef(fit), lavaan = coef(fit_ml)), 3)
```

## Practical notes

- **Order the indicators.** The other weights are ratios to the marker, so
  list first the indicator with the largest expected weight. A marker with a
  weight near zero leaves the rest poorly determined.
- **Higher-order means.** A composite's mean is fixed by the means of its
  indicators, so a latent variable measured only through composites needs its
  mean fixed at zero; INLAvaan stops with an error otherwise. For growth
  models on composites, use `asem()` with `meanstructure = TRUE` instead of
  `agrowth()`.
- **Model comparison.** As in lavaan, the covariances of each composite's
  indicators are fixed at their sample values, so marginal likelihoods and the
  DIC compare only models with the same composites.
- **Limits.** Composites are supported in single-level models with continuous
  data, in one or more groups. Two-level models, ordinal data,
  `composites.cov = "free"` and covariances between the indicators of a
  composite and other variables stop with an error.

## References
