---
title: 'Bayesian Fit Indices'
description: >
  A guide to the Bayesian fit indices available in INLAvaan, including mathematical definitions, worked examples, and comparison with blavaan.
vignette: >
  %\VignetteIndexEntry{Bayesian Fit Indices}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk: 
    collapse: true
    comment: '#>'
bibliography: ../refs.bib
toc: true
---

```{r}
#| label: setup
#| include: false
library(INLAvaan)
library(lavaan)
```

## Introduction

Classical SEM fit indices (RMSEA, CFI, TLI, etc.) are computed from a single maximum-likelihood chi-square statistic.
In Bayesian SEM, the parameters are random, so **each posterior draw** yields a different chi-square value and, therefore, a different set of fit indices.
INLAvaan provides Bayesian analogues of the most common indices, following the framework of @garnier2021bayesian (as implemented in **blavaan**).

This vignette covers:

1. Mathematical definitions of the Bayesian fit indices.
2. The two rescaling methods (`"devM"` and `"MCMC"`).
3. A worked example with the Holzinger--Swineford data.
4. Comparison with a baseline model (incremental indices).
5. Differences between INLAvaan and blavaan.


## Mathematical background

### Deviance and chi-square

Let $\boldsymbol\theta^{(s)}$ denote the $s$-th posterior draw of the model parameters ($s = 1, \dots, S$).
The per-sample deviance chi-square is
$$
  \chi^2_s = 2 \bigl[\ell_{\text{sat}} - \ell(\boldsymbol\theta^{(s)})\bigr],
$$
where $\ell_{\text{sat}}$ is the log-likelihood of the saturated model (sample moments equal model moments) and $\ell(\boldsymbol\theta^{(s)})$ is the log-likelihood evaluated at the $s$-th draw.
This equals $N \, F_{\text{ML}}(\boldsymbol\theta^{(s)})$, where $F_{\text{ML}}$ is the ML discrepancy function.


### Rescaling

INLAvaan supports two rescaling methods, controlled by the `rescale` argument:

**`"devM"` (default).** Uses the DIC-based effective number of parameters $p_D$.
$$
  d_s = \chi^2_s - p_D, \qquad
  \mathrm{df} = p - p_D, \qquad
  N_{\mathrm{adj}} = N.
$$
If $p_D$ is unreasonable ($p_D \leq 0$ or $p_D \geq p$), INLAvaan falls back to using $p_D = q$ (the number of free parameters).

**`"MCMC"`.** Uses the classical chi-square with $N - 1$ scaling.
$$
  d_s = \frac{N - 1}{N} \chi^2_s, \qquad
  \mathrm{df} = p - q, \qquad
  N_{\mathrm{adj}} = N - G,
$$
where $q$ is the number of free parameters and $G$ is the number of groups.

In both cases the per-sample noncentrality parameter is $\hat\lambda_s = \max(d_s - \mathrm{df},\, 0)$.


### Absolute fit indices

The following indices are computed at **each** posterior draw $s$, generating a posterior distribution.

**BRMSEA (Bayesian RMSEA).**
$$
  \text{BRMSEA}_s = \sqrt{\frac{\hat\lambda_s}{\mathrm{df} \cdot N_\text{adj}}} \cdot \sqrt{G}.
$$

**BGammaHat.**
$$
  \text{BGammaHat}_s = \frac{v}{v + 2\hat\lambda_s / N_\text{adj}},
$$
where $v = \sum_g p_g$ is the total number of observed variables across groups.

**Adjusted BGammaHat.**
$$
  \text{adjBGammaHat}_s = 1 - \frac{p}{\mathrm{df}} \bigl(1 - \text{BGammaHat}_s\bigr).
$$

**BMc (McDonald's centrality index).**
$$
  \text{BMc}_s = \exp\!\bigl(-\tfrac{1}{2}\hat\lambda_s / N_\text{adj}\bigr).
$$


### Incremental fit indices

Incremental indices compare the target model against a **baseline** (null) model.
Let $d_s^{(0)}$, $\mathrm{df}^{(0)}$, and $\hat\lambda_s^{(0)}$ denote the corresponding quantities for the baseline model.

**BCFI (Bayesian CFI).**
$$
  \text{BCFI}_s = 1 - \frac{\hat\lambda_s}{\hat\lambda_s^{(0)}}.
$$

**BTLI (Bayesian TLI).**
$$
  \text{BTLI}_s = \frac{d_s^{(0)} / \mathrm{df}^{(0)} - d_s / \mathrm{df}}{d_s^{(0)} / \mathrm{df}^{(0)} - 1}.
$$

**BNFI (Bayesian NFI).**
$$
  \text{BNFI}_s = \frac{d_s^{(0)} - d_s}{d_s^{(0)}}.
$$

The posterior expectations (EAP), standard deviations, quantile-based credible intervals, and modes of these distributions are reported by the `summary()` method.


## Example: three-factor CFA

We fit a three-factor CFA on the Holzinger--Swineford (1939) data.

```{r}
#| label: fit-cfa
HS.model <- "
  visual  =~ x1 + x2 + x3
  textual =~ x4 + x5 + x6
  speed   =~ x7 + x8 + x9
"

fit <- acfa(HS.model, HolzingerSwineford1939, verbose = FALSE)
```

### Scalar fit measures

`fitMeasures()` returns the posterior mean (EAP) of each Bayesian fit index alongside the marginal log-likelihood, ppp, DIC, and gradient diagnostics.

```{r}
#| label: fitmeasures
fitMeasures(fit)
```

### Posterior distributions of fit indices

To inspect the full posterior distribution of each index, use `bfit_indices()`.
This returns an S3 object of class `"bfit_indices"`.

```{r}
#| label: bfi-object
bfi <- bfit_indices(fit)
bfi
```

Calling `summary()` provides a table of posterior summaries (Mean, SD, 2.5%, 50%, 97.5%, Mode) for each index.

```{r}
#| label: bfi-summary
summary(bfi)
```

You can also access the raw per-sample vectors for custom analysis:

```{r}
#| label: bfi-hist
#| fig-cap: "Posterior distribution of BRMSEA"
hist(bfi$indices$BRMSEA, breaks = 30, main = "BRMSEA", xlab = "BRMSEA",
     col = "steelblue", border = "white")
```


## Comparison with a baseline model

Incremental indices (BCFI, BTLI, BNFI) require a **baseline model**.
A common choice is the independence (uncorrelated) model.

```{r}
#| label: fit-baseline
null.model <- "
  x1 ~~ x1
  x2 ~~ x2
  x3 ~~ x3
  x4 ~~ x4
  x5 ~~ x5
  x6 ~~ x6
  x7 ~~ x7
  x8 ~~ x8
  x9 ~~ x9
"

fit_null <- acfa(null.model, HolzingerSwineford1939, verbose = FALSE)
```

Now pass the baseline model to `fitMeasures()` or `bfit_indices()`:

```{r}
#| label: fm-incremental
fitMeasures(fit, baseline.model = fit_null)
```

```{r}
#| label: bfi-incremental
bfi_inc <- bfit_indices(fit, baseline.model = fit_null)
summary(bfi_inc)
```


## Rescaling: `"devM"` vs `"MCMC"`

The `rescale` argument controls how the chi-square and degrees of freedom are computed.
The default is `"devM"`, which subtracts the DIC-based $p_D$ from the deviance.
To use the classical $N - 1$ scaling instead, set `rescale = "MCMC"`:

```{r}
#| label: bfi-mcmc
bfi_mcmc <- bfit_indices(fit, rescale = "MCMC")
summary(bfi_mcmc)
```

The two methods will generally produce different results, especially with informative priors or when $p_D$ deviates substantially from $q$.


## Differences from blavaan {#sec-differences}

INLAvaan's Bayesian fit indices follow the same mathematical framework as **blavaan** [@merkle2021blavaan], with a few implementation differences:

| Feature | INLAvaan | blavaan |
|:--------|:---------|:--------|
| Posterior samples | INLA-based (Sobol/NORTA) | MCMC draws from Stan/JAGS |
| Rescaling methods | `"devM"`, `"MCMC"` | `"devM"`, `"MCMC"`, `"ppmc"` |
| Effective parameters ($p_D$) | DIC-based $p_D$ only | DIC-based $p_D$, LOOIC-based $p_{\text{loo}}$, or WAIC-based $p_{\text{waic}}$ |
| HPD intervals | Not currently computed | Computed via **coda** |
| Summary statistics | Mean, SD, 2.5%, 50%, 97.5%, Mode | EAP, Median, MAP, SD, HPD |
| Return class | S3 `"bfit_indices"` | S4 `"blavFitIndices"` |

Currently, INLAvaan only supports $p_D$ from the DIC (i.e., `p_dic`).
The `"ppmc"` rescaling method (which computes replicated data under the posterior predictive) is not yet available.


## References

::: {#refs}
:::
