---
title: 'Leave-One-Out Cross-Validation'
description: >
  Refit-free LOO model assessment in INLAvaan: per-subject and per-cluster
  scores, model comparison with paired standard errors, WAIC, and scoring
  submodels without refitting.
vignette: >
  %\VignetteIndexEntry{Leave-One-Out Cross-Validation}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
bibliography: ../refs.bib
toc: true
---

```{r}
#| label: setup
#| include: false
library(INLAvaan)
library(lavaan)
```

## Introduction

How well would my model predict *new* data?
Leave-one-out (LOO) cross-validation answers this by holding out one unit at a time, refitting the model to the remaining data, and scoring the held-out unit under the refitted posterior.
The total score is the expected log predictive density,
$$
  \mathrm{elpd}_{\mathrm{loo}} = \sum_{u=1}^n \log p(y_u \mid y_{-u}),
$$
which rewards models that predict well and automatically penalises overfitting -- making it a natural criterion for comparing models.

Computed naively, LOO needs $n$ refits.
MCMC-based packages such as `{blavaan}` avoid this by importance-sampling over posterior draws [@vehtari2017practical], but this still requires the full set of MCMC draws.
INLAvaan instead exploits its Laplace machinery: `loo()` approximates each case-deletion posterior by a Taylor expansion around the full-data posterior summary, so **the entire LOO is computed from a single fit** -- no refitting and no sampling.
On the Holzinger--Swineford example below this takes a fraction of a second.

Two unit types are scored, resolved automatically from the model:

- **LOSO** (leave-one-*subject*-out): single-level models, one unit per row.
- **LOCO** (leave-one-*cluster*-out): two-level models, one unit per cluster --
  the relevant predictive question is "how well would the model predict a new
  cluster?".


## How it works

Write $\ell_u(\theta) = \log p(y_u \mid \theta)$ for unit $u$'s log-likelihood contribution, with score $s_u$ and Hessian $H_u$ evaluated at the posterior summary $(\theta^*, \Sigma)$ from the fit.
`loo()` reports first- and second-order Taylor approximations of the log conditional predictive ordinate:
$$
\begin{aligned}
  \log \mathrm{CPO}_u^{(1)} &= \ell_u - \tfrac12 s_u^\top \Sigma\, s_u, \\
  \log \mathrm{CPO}_u^{(2)} &= \ell_u
    - \tfrac12 s_u^\top (\Sigma^{-1} + H_u)^{-1} s_u
    + \tfrac12 \log \lvert I + \Sigma H_u \rvert .
\end{aligned}
$$
The headline `elpd_loo` is the sum of the second-order terms, with standard error $\sqrt{n \, \widehat{\mathrm{var}}(\log \mathrm{CPO}_u)}$, and `looic` $= -2\,\mathrm{elpd}_{\mathrm{loo}}$ on the familiar information-criterion scale.
The effective number of parameters $p_{\mathrm{loo}} = \sum_u (\mathrm{lpd}_u - \log \mathrm{CPO}_u)$ uses the analogous expansion of the full-posterior pointwise predictive density $\mathrm{lpd}_u$.
The rare unit whose second-order curvature matrix is not positive definite falls back to first order (flagged in the output); a warning is raised if this happens for many units, since it suggests the Gaussian posterior summary itself is poor.


## A first example

We fit the classic three-factor CFA to the Holzinger--Swineford data.
Fitting with `meanstructure = TRUE` is recommended for LOO: otherwise unit log-likelihoods are evaluated at zero means and absolute ELPD values are biased (comparisons between models on the same data remain valid either way).

```{r}
#| label: loso-fit
HS.model <- "
  visual  =~ x1 + x2 + x3
  textual =~ x4 + x5 + x6
  speed   =~ x7 + x8 + x9
"
fit <- acfa(HS.model, HolzingerSwineford1939, meanstructure = TRUE,
            verbose = FALSE)

(res <- loo(fit))
```

The pointwise contributions are available for inspection -- useful for spotting influential observations (a large `score_norm` or unusually low `log_cpo_2` flags a unit the model predicts poorly):

```{r}
#| label: loso-perunit
head(res$per_unit)
```


## Comparing models

Because ELPD differences between models fitted to the *same* data are paired, their standard errors should be computed from the pointwise differences rather than the marginal SEs.
`compare(..., loo = TRUE)` does this automatically:

```{r}
#| label: compare
one.factor <- "g =~ x1 + x2 + x3 + x4 + x5 + x6 + x7 + x8 + x9"
fit1f <- acfa(one.factor, HolzingerSwineford1939, meanstructure = TRUE,
              verbose = FALSE)

compare(fit, fit1f, loo = TRUE)
```

Models are sorted by descending ELPD; `elpd_diff` and `se_diff` are relative to the best model.
A common heuristic is that a difference smaller than a couple of `se_diff` units is not practically meaningful [@vehtari2017practical] -- here the three-factor model is preferred by a wide margin.


## Two-level models

For models fitted with a `cluster` argument, `loo()` automatically switches to per-cluster scoring (LOCO).
Here the covariates are modelled jointly (`fixed.x = FALSE`); the next section explains what happens under lavaan's default covariate treatment.

```{r}
#| label: loco
model2l <- "
  level: 1
    fw =~ y1 + y2 + y3
    fw ~ x1 + x2 + x3
  level: 2
    fb =~ y1 + y2 + y3
    fb ~ w1 + w2
"
fit2l <- asem(model2l, Demo.twolevel, cluster = "cluster",
              meanstructure = TRUE, fixed.x = FALSE, verbose = FALSE)

loo(fit2l)
```


## Exogenous covariates: joint and conditional scores

When a model contains exogenous covariates, the *flavour* of the LOO score follows the likelihood the model was fitted with.
Under `fixed.x = FALSE` the covariates receive a saturated Gaussian block and each unit is scored by the joint predictive density of its outcomes *and* covariates ("how surprised is the model by a brand-new unit, characteristics included?").
Under `fixed.x = TRUE` -- lavaan's default -- the fitted likelihood is the conditional one, and each unit is scored by the predictive density of its outcomes *given* its covariates ("given a new unit with known characteristics, how well do we predict its outcomes?"), the familiar regression cross-validation convention.
No refitting trickery is involved: the conditional likelihood is exactly invariant to the fixed covariate moments, so the conditional score carries no additional approximation.

```{r}
#| label: flavours
model_x <- "
  visual  =~ x1 + x2 + x3
  textual =~ x4 + x5 + x6
  speed   =~ x7 + x8 + x9
  visual ~ ageyr + grade
  textual ~ ageyr + grade
"
dat_x <- na.omit(
  HolzingerSwineford1939[, c(paste0("x", 1:9), "ageyr", "grade")]
)

# lavaan's default fixed.x = TRUE: scored conditionally on the covariates
fit_cond <- asem(model_x, dat_x, meanstructure = TRUE, verbose = FALSE)
loo(fit_cond)
```

The two flavours estimate different quantities whose scales differ by the covariate predictive density, so **a joint and a conditional elpd must never appear in the same comparison** -- `compare(..., loo = TRUE)` refuses mixed-flavour comparisons outright.
Within the conditional flavour, however, models conditioning on *different* covariate sets are directly comparable as long as the outcome variables match, which makes covariate selection straightforward:

```{r}
#| label: covariate-selection
model_x1 <- "
  visual  =~ x1 + x2 + x3
  textual =~ x4 + x5 + x6
  speed   =~ x7 + x8 + x9
  visual ~ ageyr
"
fit_cond1 <- asem(model_x1, dat_x, meanstructure = TRUE, verbose = FALSE)

compare(fit_cond, fit_cond1, loo = TRUE)
```

(Under the joint flavour the same comparison would require retaining `grade` in both models, since joint scores of models spanning different variable sets live on different sample spaces.)
Both flavours support any covariate placement, including cluster-level and within-level covariates in two-level models.

## Storing the result with the fit

`loo(fit)` never modifies the fitted object -- but under the default
`test = "standard"`, the fit itself already computes and stores both the full LOO and the WAIC whenever the model is supported, has a mean structure, and (for LOO) the predicted serial cost is within a 10-second budget.
The prediction is calibrated at run time by timing a single score evaluation, so on typical single-level models -- where the full LOO costs a fraction of the fit itself -- you simply get `loo(fit)` and `waic(fit)` for free.
The WAIC reuses the very draws the fit produced for its posterior summaries, so it costs only one casewise pass.

For expensive cases you can force the LOO regardless of the budget at fit time,

```{r}
#| label: fit-time
#| eval: false
fit <- acfa(HS.model, HolzingerSwineford1939, meanstructure = TRUE,
            test = c("standard", "loo"))
```

or store it afterwards with an explicit reassignment:

```{r}
#| label: add-loo
fit <- add_loo(fit)
```

A stored result is returned instantly by `loo(fit)` and reused by `fitMeasures()` (where the blavaan-style names appear) and `compare(..., loo = TRUE)`:

```{r}
#| label: fitmeasures
fitMeasures(fit, c("elpd_loo", "se_loo", "p_loo", "looic"))
```


## WAIC

The widely applicable information criterion [@watanabe2010asymptotic] is asymptotically equivalent to LOO and is also available.
Unlike `loo()`, `waic()` is sampling-based: it evaluates unit log-likelihoods over posterior draws, so results carry Monte Carlo error, and units with $p_{\mathrm{waic},u} > 0.4$ trigger a reliability warning -- in which case prefer `loo()` [@vehtari2017practical].

```{r}
#| label: waic
#| warning: false
set.seed(1)
waic(fit)
```


## Scoring submodels without refitting

The `theta` and `Sigma` arguments evaluate the LOO at an *arbitrary* Gaussian posterior summary instead of the fit's own.
Combined with Gaussian conditioning, this scores a constrained submodel from the encompassing fit alone.
For example, to score the submodel with the `visual ~~ speed` covariance fixed to zero, condition the summary on that parameter and re-evaluate:

```{r}
#| label: override
int <- get_inlavaan_internal(fit)
theta <- int$theta_star
Sigma <- int$Sigma_theta

p <- which(names(coef(fit)) == "visual~~speed")
theta_c <- theta - Sigma[, p] * (theta[p] / Sigma[p, p])
Sigma_c <- Sigma - tcrossprod(Sigma[, p]) / Sigma[p, p]

loo(fit, theta = theta_c, Sigma = Sigma_c)
```

No refit took place: the conditioned summary has the covariance locked at zero (its row and column of `Sigma_c` vanish), and the LOO machinery automatically restricts to the remaining parameters.
This pair of arguments is the building block for custom model-search strategies -- screen many candidate restrictions by conditioning, score each with `loo()`, and only refit the winners.
INLAvaan deliberately provides just this evaluation API; the search logic is yours to design.


## Practical notes

- **Supported models.** Continuous-indicator models fitted with the `ML`
  estimator, single-group or multigroup -- groups are
  independent, so each unit is scored against its own group's moments,
  with a `group` column in the pointwise table (see the
  [multigroup article](multigroup.html) for the measurement-invariance
  workflow). Fits with missing data are scored under full-information
  maximum likelihood (`missing = "ml"`), single- or two-level; see the
  [missing-data article](missing.html). Ordinal (PML) and multigroup
  two-level models are not supported yet. Models with exogenous
  covariates are scored jointly (`fixed.x = FALSE`) or conditionally
  (`fixed.x = TRUE`), following the fitted likelihood, for any covariate
  placement.
- **Missing data.** Under FIML each unit is scored on the entries it
  actually has -- the observed-data predictive, with the full row
  (single-level) or whole cluster (two-level LOCO) deleted from the
  conditioning set -- carrying the same missing-at-random assumption as the
  fit. A single-level unit with fewer observed entries self-weights,
  contributing a smaller score; a two-level cluster contributes its
  observed-data marginal likelihood. Two missing-data fits are comparable
  with `compare()` only when they share the same observed entries (the same
  data *and* the same holes). The two-level conditional predictive
  (`type = "loso"`) is available under missing data too.
- **Parallelism is opt-in.** The default runs serially; pass
  `loo(fit, cores = 2)` to parallelise the Hessian stage via forking
  (not available on Windows).
- **First order vs second order.** The second-order correction matters: in
  our validation it tracks brute-force refits to within about one ELPD unit,
  while the first-order score can be off by tens of units. Use
  `second_order = FALSE` only for quick screening.
- **Marginal vs conditional predictive on two-level models.** The default
  `type = "loco"` is the *marginal* predictive (leave-one-cluster-out:
  prediction for a *new* cluster). `loo(fit2l, type = "loso")` -- and
  `waic(fit2l, type = "loso")` -- instead score the *conditional*
  predictive (leave-one-unit-out: a new observation within an *observed*
  cluster, each contribution the conditional density of a row given the
  rest of its cluster). These answer different questions and are easily
  conflated [@merkle2019bayesian], so the marginal is the default and the
  conditional warns. It is available with and without missing data and is
  expensive for large datasets -- subset with `units`.

## References
