---
title: 'Leave-One-Out Cross-Validation'
description: >
  Deterministic LOO model assessment in INLAvaan via per-subject and per-cluster
  scores, model comparison with paired standard errors, WAIC, and scoring
  submodels without refitting.
vignette: >
  %\VignetteIndexEntry{Leave-One-Out Cross-Validation}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
bibliography: ../refs.bib
toc: true
---

```{r}
#| label: setup
#| include: false
library(INLAvaan)
library(lavaan)
```

## Introduction

How well would my model predict *new* data?
Leave-one-out (LOO) cross-validation answers this by holding out one unit at a time, refitting the model to the remaining data, and scoring the held-out unit under the refitted posterior.
The total score is the expected log predictive density,
$$
  \mathrm{elpd}_{\mathrm{loo}} = \sum_{u=1}^n \log p(y_u \mid y_{-u}),
$$
which rewards models that predict well and automatically penalises overfitting, making it a natural criterion for comparing models.

Computed naively, LOO needs $n$ refits.
MCMC-based packages such as `{blavaan}` avoid this by importance-sampling over posterior draws [@vehtari2017practical], but this still requires the full set of MCMC draws.
INLAvaan instead exploits its Laplace machinery: `loo()` approximates each case-deletion posterior by a Taylor expansion around the full-data posterior summary, so **the entire LOO is computed from a single fit**, with no refitting and no sampling.

Two unit types are scored, resolved automatically from the model:

- **LOSO** (leave-one-*subject*-out): single-level models, one unit per row.
- **LOCO** (leave-one-*cluster*-out): two-level models, one unit per cluster --
  the relevant predictive question is "how well would the model predict a new
  cluster?".


## Technical details in brief

`loo()` computes everything from the fit's Laplace summary $\mathcal{N}(\theta^*, \Omega)$.
An exact identity turns each held-out predictive density into an expectation under the *full-data* posterior, evaluated in closed form by Taylor-expanding the unit's log-likelihood $\ell_u(\theta) = \log p(y_u \mid \theta)$ about $\theta^*$.
With $s_u$ and $H_u$ the gradient and Hessian of $\ell_u$ there, the two orders are
$$
\begin{aligned}
  \log \mathrm{CPO}_u^{(1)} &= \ell_u - \tfrac12 s_u^\top \Omega\, s_u, \\
  \log \mathrm{CPO}_u^{(2)} &= \ell_u
    - \tfrac12 s_u^\top (\Omega^{-1} + H_u)^{-1} s_u
    + \tfrac12 \log \lvert I + \Omega H_u \rvert ,
\end{aligned}
$$
the second order additionally handing back the information the unit itself lent to the posterior.
The headline `elpd_loo` is the sum of the second-order terms, with standard error $\sqrt{n \, \widehat{\mathrm{var}}(\log \mathrm{CPO}_u)}$; `looic` $= -2\,\mathrm{elpd}_{\mathrm{loo}}$, and `p_loo` (the **loo** package's effective number of parameters, matching `loo::loo()`) comes from the analogous expansion of the pointwise predictive density.

Two checks, both free with the result, tell you whether to trust the expansion.
The first is an existence condition and is enforced automatically: $\log \mathrm{CPO}_u^{(2)}$ exists exactly when $\Omega^{-1} + H_u \succ 0$, equivalently when the unit's *leverage* `k_max` (the largest eigenvalue of $-\Omega H_u$, reported in `per_unit`) stays below 1.
A unit at or above 1 carries as much curvature as the whole posterior in some direction, so its true LOO term genuinely is extreme.
In such cases, the whole result then reverts to first order, with a warning naming the units to inspect.

The second is a gap check, printed with every second-order result.
The first- and second-order totals of a settled expansion differ by $p_D/2$, approached *from above*, so the printout reports the gap, the reference $p_D/2$ (`pd_trace`, the trace form $\operatorname{tr}(-\Omega \sum_u H_u)$), and the excess of one over the other.
A large positive excess says the expansion has not settled over the sample.
No threshold is applied, and the reading is left to the user, but in our validation the gap was always within a few elpd units of zero for well-behaved models.


## A first example

We fit the classic three-factor CFA to the Holzinger--Swineford data.

```{r}
#| label: loso-fit
HS.model <- "
  visual  =~ x1 + x2 + x3
  textual =~ x4 + x5 + x6
  speed   =~ x7 + x8 + x9
"
fit <- acfa(HS.model, HolzingerSwineford1939, meanstructure = TRUE,
            verbose = FALSE)

(res <- loo(fit))
```

The printout ends with the gap check described above.
The pointwise contributions are available for inspection as well.
E.g. a large `score_norm` or an unusually low `log_cpo_2` flags a unit the model predicts poorly, and `k_max` is the unit's leverage:

```{r}
#| label: loso-perunit
head(res$per_unit)
```

Because elpd differences between models fitted to the *same* data are paired, their standard errors should be computed from the pointwise differences rather than the marginal SEs.
`compare(..., loo = TRUE)` does this automatically:

```{r}
#| label: compare
one.factor <- "g =~ x1 + x2 + x3 + x4 + x5 + x6 + x7 + x8 + x9"
fit1f <- acfa(one.factor, HolzingerSwineford1939, meanstructure = TRUE,
              verbose = FALSE)

compare(fit, fit1f, loo = TRUE)
```

Models are sorted by descending elpd, and `elpd_diff` and `se_diff` are relative to the best model.
A common heuristic is that a difference smaller than a couple of `se_diff` units is not practically meaningful [@vehtari2017practical].
Here the three-factor model is preferred by a wide margin.


## Two-level models

For models fitted with a `cluster` argument, `loo()` automatically switches to per-cluster scoring (LOCO).
Here the covariates are modelled jointly (`fixed.x = FALSE`) though the default `fixed.x = TRUE` is also supported.
See the practical considerations below for a discussion of the two flavours.

```{r}
#| label: loco
model2l <- "
  level: 1
    fw =~ y1 + y2 + y3
    fw ~ x1 + x2 + x3
  level: 2
    fb =~ y1 + y2 + y3
    fb ~ w1 + w2
"
fit2l <- asem(model2l, Demo.twolevel, cluster = "cluster",
              meanstructure = TRUE, fixed.x = FALSE, verbose = FALSE)
(loco <- loo(fit2l))  # type = "loco" is automatic
```

With the per unit dataframe stored, we can perform additional diagnostics, such as to query the relationship between the second-order log predictive density and the second-order log CPO.
A unit sitting above the diagonal in the plot below is one the full-data posterior flatters, since its in-sample density exceeds its held-out one.
The vertical gap is exactly its contribution to `p_loo`.

```{r}
#| label: loco-perunit
plot(lpd_2 ~ log_cpo_2, loco$per_unit)
```


There is no need to score every unit each time.
The `units` argument restricts the pass to a chosen subset, given as cluster positions for LOCO and case numbers for LOSO, and the reported totals then cover that subset alone.
This is useful when a full pass is expensive, or to revisit the units the diagnostics above single out.
Here we rescore the three worst-predicted clusters:

```{r}
#| label: loco-units
worst <- with(loco$per_unit, unit[order(log_cpo_2)][1:3])
loo(fit2l, units = worst)
```


## Storing the result with the fit

`loo(fit)` never modifies the fitted object -- but under the default
`test = "standard"`, the fit itself already computes and stores both the full LOO and the WAIC whenever the model is supported, has a mean structure, and (for LOO) the predicted serial cost is within a 10-second budget.
The prediction is calibrated at run time by timing a single score evaluation, so on typical single-level models -- where the full LOO costs a fraction of the fit itself -- you simply get `loo(fit)` and `waic(fit)` for free.
The WAIC reuses the very draws the fit produced for its posterior summaries, so it costs only one casewise pass.

For expensive cases you can force the LOO regardless of the budget at fit time,

```{r}
#| label: fit-time
#| eval: false
fit <- acfa(HS.model, HolzingerSwineford1939, meanstructure = TRUE,
            test = c("standard", "loo"))
```

or store it afterwards with an explicit reassignment:

```{r}
#| label: add-loo
fit <- add_loo(fit)
```

A stored result is returned instantly by `loo(fit)` and reused by `fitmeasures()` (where the blavaan-style names appear) and `compare(..., loo = TRUE)`:

```{r}
#| label: fitmeasures
fitmeasures(fit, c("elpd_loo", "se_loo", "p_loo", "looic"))
```

<!-- 
## WAIC

The widely applicable information criterion [@watanabe2010asymptotic] is asymptotically equivalent to LOO and is also available.
Like `loo()`, `waic()` is computed in closed form from the Laplace summary -- the penalty is the variance of the unit log-likelihood under the Gaussian posterior, a polynomial in its moments -- so it involves no posterior draws and carries no Monte Carlo error.
At first order the two are identical ($\mathrm{lpd}^{(1)}_u - p^{(1)}_{\mathrm{waic},u} = \log \mathrm{CPO}^{(1)}_u$ exactly); at second order they differ by a curvature gap, with WAIC weakly optimistic relative to LOO.
No reliability threshold is applied to $p_{\mathrm{waic}}$: being a polynomial in the posterior moments it is always finite, so the only condition the WAIC carries is the one its $\mathrm{lpd}$ term inherits -- $\Omega^{-1} - H_u \succ 0$, equivalently $k_{\min} > -1$ -- and a unit failing it sends every estimate to first order, with a warning.

```{r}
#| label: waic
waic(fit)
```
 -->

## Scoring submodels without refitting

The `theta` and `Omega` arguments evaluate the LOO at an *arbitrary* Gaussian posterior summary $(\theta^*, \Omega)$ instead of the fit's own.
Combined with Gaussian conditioning, this scores a constrained submodel from the encompassing fit alone.
For example, to score the submodel with the `visual ~~ speed` covariance fixed to zero, condition the summary on that parameter and re-evaluate:

```{r}
#| label: override
int <- get_inlavaan_internal(fit)
theta <- int$theta_star
Omega <- int$Sigma_theta

p <- which(names(coef(fit)) == "visual~~speed")
theta_c <- theta - Omega[, p] * (theta[p] / Omega[p, p])
Omega_c <- Omega - tcrossprod(Omega[, p]) / Omega[p, p]

loo(fit, theta = theta_c, Omega = Omega_c)
```

No refit took place: the conditioned summary has the covariance locked at zero (its row and column of `Omega_c` vanish), and the LOO machinery automatically restricts to the remaining parameters.
This pair of arguments is the building block for custom model-search strategies -- screen many candidate restrictions by conditioning, score each with `loo()`, and only refit the winners.
INLAvaan deliberately provides just this evaluation API; the search logic is yours to design.


## Practical considerations

- **Compare models at second order only** (the default). The first-order
  score overstates the elpd by $p_D/2$, which grows with model dimension;
  in our validation the second-order score tracks brute-force refits to
  within about one elpd unit, while first order can be off by tens.
  `compare()` scores every model at one common order and states it.
- **Joint and conditional scores never mix.** With exogenous covariates the
  score follows the fitted likelihood: under `fixed.x = FALSE` a unit is
  scored by the joint predictive density of its outcomes *and* covariates,
  while under `fixed.x = TRUE` (lavaan's default) it is scored by the
  density of its outcomes *given* its covariates, the familiar regression
  cross-validation convention. The two estimate different quantities, so
  `compare(..., loo = TRUE)` refuses to put them in one table. Within the
  conditional flavour, models with *different* covariate sets are directly
  comparable as long as the outcomes match (the covariate-selection
  setting); joint scores require identical variable sets across models.
- **Fit with `meanstructure = TRUE`**, otherwise absolute elpd values are
  biased (same-data comparisons remain valid either way).
- **LOCO and LOSO answer different questions.** On two-level fits the
  default `type = "loco"` is the marginal predictive (a *new* cluster);
  `type = "loso"` scores the conditional one (a new row in an *observed*
  cluster) and warns, since the two are easily conflated
  [@merkle2019bayesian].
- **Missing data.** FIML fits are scored on the observed entries, so two
  such fits are comparable only when they share the same data *and* the same
  holes; see the [missing-data article](missing.html).
- **Scope and cost.** Supported are continuous-indicator `ML` fits, single-
  or two-level, single-group or multigroup (not ordinal PML, not multigroup
  two-level). Parallelism is opt-in via `cores`, and `units` scores a
  subset.

## References
