---
title: 'Leave-One-Out Cross-Validation'
description: >
  Refit-free LOO model assessment in INLAvaan: per-subject and per-cluster
  scores, model comparison with paired standard errors, WAIC, and scoring
  submodels without refitting.
vignette: >
  %\VignetteIndexEntry{Leave-One-Out Cross-Validation}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
bibliography: ../refs.bib
toc: true
---

```{r}
#| label: setup
#| include: false
library(INLAvaan)
library(lavaan)
```

## Introduction

How well would my model predict *new* data?
Leave-one-out (LOO) cross-validation answers this by holding out one unit at a time, refitting the model to the remaining data, and scoring the held-out unit under the refitted posterior.
The total score is the expected log predictive density,
$$
  \mathrm{elpd}_{\mathrm{loo}} = \sum_{u=1}^n \log p(y_u \mid y_{-u}),
$$
which rewards models that predict well and automatically penalises overfitting -- making it a natural criterion for comparing models.

Computed naively, LOO needs $n$ refits.
MCMC-based packages such as `{blavaan}` avoid this by importance-sampling over posterior draws [@vehtari2017practical], but this still requires the full set of MCMC draws.
INLAvaan instead exploits its Laplace machinery: `loo()` approximates each case-deletion posterior by a Taylor expansion around the full-data posterior summary, so **the entire LOO is computed from a single fit** -- no refitting and no sampling.
On the Holzinger--Swineford example below this takes a fraction of a second.

Two unit types are scored, resolved automatically from the model:

- **LOSO** (leave-one-*subject*-out): single-level models, one unit per row.
- **LOCO** (leave-one-*cluster*-out): two-level models, one unit per cluster --
  the relevant predictive question is "how well would the model predict a new
  cluster?".


## How it works

Write $\ell_u(\theta) = \log p(y_u \mid \theta)$ for unit $u$'s log-likelihood contribution, with score $s_u$ and Hessian $H_u$ evaluated at the posterior summary $(\theta^*, \Sigma)$ from the fit.
`loo()` reports first- and second-order Taylor approximations of the log conditional predictive ordinate:
$$
\begin{aligned}
  \log \mathrm{CPO}_u^{(1)} &= \ell_u - \tfrac12 s_u^\top \Sigma\, s_u, \\
  \log \mathrm{CPO}_u^{(2)} &= \ell_u
    - \tfrac12 s_u^\top (\Sigma^{-1} + H_u)^{-1} s_u
    + \tfrac12 \log \lvert I + \Sigma H_u \rvert .
\end{aligned}
$$
The headline `elpd_loo` is the sum of the second-order terms, with standard error $\sqrt{n \, \widehat{\mathrm{var}}(\log \mathrm{CPO}_u)}$, and `looic` $= -2\,\mathrm{elpd}_{\mathrm{loo}}$ on the familiar information-criterion scale.
The effective number of parameters $p_{\mathrm{loo}} = \sum_u (\mathrm{lpd}_u - \log \mathrm{CPO}_u)$ uses the analogous expansion of the full-posterior pointwise predictive density $\mathrm{lpd}_u$.
The rare unit whose second-order curvature matrix is not positive definite falls back to first order (flagged in the output); a warning is raised if this happens for many units, since it suggests the Gaussian posterior summary itself is poor.


## A first example

We fit the classic three-factor CFA to the Holzinger--Swineford data.
Fitting with `meanstructure = TRUE` is recommended for LOO: otherwise unit log-likelihoods are evaluated at zero means and absolute ELPD values are biased (comparisons between models on the same data remain valid either way).

```{r}
#| label: loso-fit
HS.model <- "
  visual  =~ x1 + x2 + x3
  textual =~ x4 + x5 + x6
  speed   =~ x7 + x8 + x9
"
fit <- acfa(HS.model, HolzingerSwineford1939, meanstructure = TRUE,
            verbose = FALSE)

(res <- loo(fit))
```

The pointwise contributions are available for inspection -- useful for spotting influential observations (a large `score_norm` or unusually low `log_cpo_2` flags a unit the model predicts poorly):

```{r}
#| label: loso-perunit
head(res$per_unit)
```


## Comparing models

Because ELPD differences between models fitted to the *same* data are paired, their standard errors should be computed from the pointwise differences rather than the marginal SEs.
`compare(..., loo = TRUE)` does this automatically:

```{r}
#| label: compare
one.factor <- "g =~ x1 + x2 + x3 + x4 + x5 + x6 + x7 + x8 + x9"
fit1f <- acfa(one.factor, HolzingerSwineford1939, meanstructure = TRUE,
              verbose = FALSE)

compare(fit, fit1f, loo = TRUE)
```

Models are sorted by descending ELPD; `elpd_diff` and `se_diff` are relative to the best model.
A common heuristic is that a difference smaller than a couple of `se_diff` units is not practically meaningful [@vehtari2017practical] -- here the three-factor model is preferred by a wide margin.


## Two-level models

For models fitted with a `cluster` argument, `loo()` automatically switches to per-cluster scoring (LOCO).
Note that models with exogenous covariates must be fitted with `fixed.x = FALSE`, so that the covariates are part of the joint model being cross-validated.

```{r}
#| label: loco
model2l <- "
  level: 1
    fw =~ y1 + y2 + y3
    fw ~ x1 + x2 + x3
  level: 2
    fb =~ y1 + y2 + y3
    fb ~ w1 + w2
"
fit2l <- asem(model2l, Demo.twolevel, cluster = "cluster",
              meanstructure = TRUE, fixed.x = FALSE, verbose = FALSE)

loo(fit2l)
```


## Storing the result with the fit

`loo(fit)` never modifies the fitted object.
To compute the LOO once and keep it with the fit, either request it at fit time,

```{r}
#| label: fit-time
#| eval: false
fit <- acfa(HS.model, HolzingerSwineford1939, meanstructure = TRUE,
            test = c("standard", "loo"))
```

or store it afterwards with an explicit reassignment:

```{r}
#| label: add-loo
fit <- add_loo(fit)
```

A stored result is returned instantly by `loo(fit)` and reused by `fitMeasures()` (where the blavaan-style names appear) and `compare(..., loo = TRUE)`:

```{r}
#| label: fitmeasures
fitMeasures(fit, c("elpd_loo", "se_loo", "p_loo", "looic"))
```


## WAIC

The widely applicable information criterion [@watanabe2010asymptotic] is asymptotically equivalent to LOO and is also available.
Unlike `loo()`, `waic()` is sampling-based: it evaluates unit log-likelihoods over posterior draws, so results carry Monte Carlo error, and units with $p_{\mathrm{waic},u} > 0.4$ trigger a reliability warning -- in which case prefer `loo()` [@vehtari2017practical].

```{r}
#| label: waic
#| warning: false
set.seed(1)
waic(fit)
```


## Scoring submodels without refitting

The `theta` and `Sigma` arguments evaluate the LOO at an *arbitrary* Gaussian posterior summary instead of the fit's own.
Combined with Gaussian conditioning, this scores a constrained submodel from the encompassing fit alone.
For example, to score the submodel with the `visual ~~ speed` covariance fixed to zero, condition the summary on that parameter and re-evaluate:

```{r}
#| label: override
int <- get_inlavaan_internal(fit)
theta <- int$theta_star
Sigma <- int$Sigma_theta

p <- which(names(coef(fit)) == "visual~~speed")
theta_c <- theta - Sigma[, p] * (theta[p] / Sigma[p, p])
Sigma_c <- Sigma - tcrossprod(Sigma[, p]) / Sigma[p, p]

loo(fit, theta = theta_c, Sigma = Sigma_c)
```

No refit took place: the conditioned summary has the covariance locked at zero (its row and column of `Sigma_c` vanish), and the LOO machinery automatically restricts to the remaining parameters.
This pair of arguments is the building block for custom model-search strategies -- screen many candidate restrictions by conditioning, score each with `loo()`, and only refit the winners.
INLAvaan deliberately provides just this evaluation API; the search logic is yours to design.


## Practical notes

- **Supported models.** Single-group, complete-data, continuous-indicator
  models fitted with the `ML` estimator. Multigroup, missing-data, and
  ordinal (PML) models are not supported yet; models with exogenous
  covariates require `fixed.x = FALSE`.
- **Parallelism is opt-in.** The default runs serially; pass
  `loo(fit, cores = 2)` to parallelise the Hessian stage via forking
  (not available on Windows).
- **First order vs second order.** The second-order correction matters: in
  our validation it tracks brute-force refits to within about one ELPD unit,
  while the first-order score can be off by tens of units. Use
  `second_order = FALSE` only for quick screening.
- **Row deletion on two-level models.** `loo(fit2l, type = "loso")` forces
  per-row scoring on a two-level fit (each contribution is the conditional
  density of a row given the rest of its cluster). It is provided as a
  diagnostic, warns when used, and is expensive for large datasets --
  consider the `units` argument to subset.

## References
