---
title: 'Mean structures'
description: >
  How INLAvaan treats the mean structure: modelled, marginalised, or fixed,
  and when model comparisons across these choices are meaningful.
vignette: >
  %\VignetteIndexEntry{Mean structures}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
toc: true
---

```{r}
#| label: setup
#| include: false
library(INLAvaan)
library(lavaan)
set.seed(2026)
```

## Three intents, three models

A lavaan model specification expresses one of three intents about the
observed-variable means, and INLAvaan gives each a distinct Bayesian
treatment:

| You write | Meaning | Treatment |
|---|---|---|
| `meanstructure = TRUE` | means are modelled | intercepts $\boldsymbol\nu$ get priors and are estimated |
| `meanstructure = FALSE` | means are not of interest | saturated means get flat priors and are **marginalised analytically** |
| `x1 ~ 0*1` etc. | means are truly zero | fixed at zero, no parameters |

The middle row deserves explanation. Classically, `meanstructure = FALSE`
fits the covariance structure only: the likelihood is *profiled*, with each
mean plugged at its sample value. Profiling is not a Bayesian operation —
a posterior built from a profiled likelihood has no generative
interpretation. INLAvaan instead reads the same model specification as
*saturated means with flat priors, integrated out*. The integral is
available in closed form: with
$\mathbf{A} = \sum_i (\mathbf{y}_i - \bar{\mathbf{y}})(\mathbf{y}_i -
\bar{\mathbf{y}})'$,

$$
\log p(\mathbf{Y} \mid \boldsymbol\Sigma)
  = -\frac{n-1}{2}\log|\boldsymbol\Sigma|
    - \frac{1}{2}\operatorname{tr}\left(\boldsymbol\Sigma^{-1}\mathbf{A}\right)
    - \frac{p}{2}\log n + \text{const},
$$

the profiled likelihood with $n-1$ in place of $n$. This is the same
treatment R-INLA applies to any included fixed effect: its default
intercept prior is exactly flat (`prec.intercept = 0`), and the intercept
is integrated out analytically as part of the latent Gaussian field.
Relative to the profiled fit, posterior modes of variance parameters
recalibrate by the factor $n/(n-1)$ — a small, principled shift.

Note that `meanstructure = FALSE` is *not* a shortcut for fixing
intercepts at zero. Zero intercepts are a falsifiable restriction: the
sample means enter the likelihood and are tested against zero, which on
uncentred data badly distorts the covariance fit. `meanstructure = FALSE`
discards the sample means entirely; the covariance estimates are identical
to those from a model with freely estimated intercepts.

Two practical notes. Two-level models require a mean structure (the
between-level statistics *are* the cluster means), so an explicit
`meanstructure = FALSE` with `cluster =` warns and fits with
`meanstructure = TRUE`. With missing data, `missing = "ML"` likewise
implies a mean structure.

## Predictive quantities

Under the marginalised reading the saturated means are not discarded —
they have an exact posterior, $\boldsymbol\nu \mid \boldsymbol\Sigma,
\mathbf{Y} \sim N(\bar{\mathbf{y}}, \boldsymbol\Sigma/n)$. Everything
downstream conditions on it or integrates it: `loo()` and `waic()` score
each unit by the exchangeable leave-one-out conditional
$$
\ell_i = \log \varphi\left(\mathbf{y}_i;\, \bar{\mathbf{y}}_{-i},\,
  \tfrac{n}{n-1}\,\boldsymbol\Sigma\right),
$$
the exact case-deletion predictive (note the mean-estimation discount
built into both arguments); factor scores condition on
$\bar{\mathbf{y}}$; and posterior predictive replicates from `sampling()`
and `simulate()` draw the means from their posterior, inflating the
predictive covariance to $(1 + \tfrac1n)\boldsymbol\Sigma$ exactly as
estimated intercepts would.

## Worked example: a measurement invariance ladder

The two mean treatments meet in measurement invariance testing. The first
two rungs are covariance-only; the third constrains intercepts across
groups, which *forces* modelled means — there is no "marginalised scalar
invariance", because constrained means are no longer saturated.

```{r}
#| label: mi-ladder
#| message: false
model <- "
  visual  =~ x1 + x2 + x3
  textual =~ x4 + x5 + x6
  speed   =~ x7 + x8 + x9
"
dat <- lavaan::HolzingerSwineford1939

fit_configural <- acfa(model, dat, group = "school",
                       meanstructure = FALSE, verbose = FALSE)
fit_metric <- acfa(model, dat, group = "school",
                   meanstructure = FALSE, group.equal = "loadings",
                   verbose = FALSE)
fit_scalar <- acfa(model, dat, group = "school",
                   group.equal = c("loadings", "intercepts"),
                   verbose = FALSE)
```

## When are these models comparable?

Within the covariance-only rungs, marginal log-likelihoods and Bayes
factors are fine: both models carry the same flat-prior mean treatment,
so its arbitrary normalisation cancels.

```{r}
#| label: compare-within
compare(fit_configural, fit_metric)
```

Across the metric–scalar boundary they are **not**: an improper flat
prior is only defined up to a constant $c$, the marginal likelihood
scales with $c^{\,p}$ per group, and against a model with proper-prior
means that constant is orphaned — the "Bayes factor" becomes an artefact
of a normalisation convention (and even proper-but-vague intercept priors
leave it hostage to the prior width, the Lindley–Bartlett phenomenon).
`compare()` warns when fits mix the two mean treatments.

The leave-one-out conditionals, by contrast, are proper under both
treatments — $p(\mathbf{y}_i \mid \mathbf{Y}_{-i}) =
p(\mathbf{Y})/p(\mathbf{Y}_{-i})$ cancels the flat-prior constant — so
ELPD comparisons are valid across the boundary. For single-group models,
`compare(..., loo = TRUE)` is the honest cross-treatment instrument (it
still warns you to read only the ELPD columns, since the marginal
log-likelihood columns remain in the table); multigroup LOO support is
planned.

```{r}
#| label: loo-single
#| message: false
fit_f <- acfa(model, dat, meanstructure = FALSE, verbose = FALSE)
fit_t <- acfa(model, dat, meanstructure = TRUE, verbose = FALSE)
compare(fit_f, fit_t, loo = TRUE)
```

The two ELPDs agree closely: both estimate
$\sum_i \log p(\mathbf{y}_i \mid \mathbf{Y}_{-i})$, and the models differ
only in whether the means are estimated under proper priors or
marginalised under flat ones.
