---
title: 'Missing Values'
# descriotion:
vignette: >
  %\VignetteIndexEntry{Missing Values}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk: 
    collapse: true
    comment: '#>'
---

`{INLAvaan}` handles missing data in one of two ways: listwise deletion (default, i.e. uses all complete cases) or Full Information Maximum Likelihood (FIML; `missing = "ML"`).

## Simulate Data

```{r}
library(INLAvaan)
mod <- "
    # Latent variable definitions
    ind60 =~ x1 + x2 + x3
    dem60 =~ y1 + y2 + y3 + y4
    dem65 =~ y5 + y6 + y7 + y8

    # Latent regressions
    dem60 ~ ind60
    dem65 ~ ind60 + dem60

    # Residual correlations
    y1 ~~ y5
    y2 ~~ y4 + y6
    y3 ~~ y7
    y4 ~~ y8
    y6 ~~ y8
  "
dat <- lavaan::PoliticalDemocracy

# Simulate missingness (MCAR)
set.seed(221)
mis <- matrix(rbinom(prod(dim(dat)), 1, 0.99), nrow(dat), ncol(dat))
datmiss <- dat * mis
datmiss[datmiss == 0] <- NA
```

## Listwise Deletion

```{r}
fit1 <- asem(mod, datmiss, meanstructure = TRUE)
fit1@Data@nobs[[1]] == nrow(datmiss[complete.cases(datmiss), ])
print(fit1)
coef(fit1)
```

## Full Information Maximum Likelihood (FIML)

```{r}
fit2 <- asem(mod, datmiss, missing = "ML", meanstructure = TRUE)
print(fit2)
coef(fit2)
```

```{r}
plot(
  coef(fit1),
  coef(fit2),
  xlab = "Listwise Deletion Estimates",
  ylab = "FIML Estimates"
)
abline(0, 1)
```

## Model criteria under FIML

`loo()` and `waic()` work directly on a FIML fit. Each unit is scored on the entries it actually has -- the observed-data predictive $\log p(y_{i,\text{obs}} \mid D_{-i})$, with the full row deleted from the conditioning set -- so a case with more missing entries contributes a smaller log-likelihood term *and* a smaller score, self-weighting in the expected log predictive density. The missing-at-random assumption that justifies FIML estimation also justifies this predictive score.

```{r}
loo(fit2)
```

Comparing two FIML fits with `compare(..., loo = TRUE)` is valid only when they share the same observed entries -- the same data *and* the same missingness pattern -- since each unit is scored on the entries it has. See the [cross-validation article](loo.html) for the Taylor case-deletion method itself.


