---
title: 'Multilevel SEM'
# descriotion:
vignette: >
  %\VignetteIndexEntry{Two-level SEM}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk: 
    collapse: true
    comment: '#>'
bibliography: ../refs.bib
---

Standard SEM assumes that all observations are independent. 
However, data often have a nested structure (e.g., patients within hospitals, employees within companies, or students within schools). 
Ignoring this structure assumes independence, leading to underestimated posterior uncertainty (overconfidence) and potentially biased parameter estimates.

This vignette demonstrates how to estimate a two-level SEM using INLAvaan.
A multilevel SEM decomposes the covariance matrix into separate levels:

1. **Within-level:** Variation among individuals relative to their group mean.
2. **Between-level:** Variation of the group means themselves.

This allows us to ask distinct questions at each level. 
For example:

> *"Does a student's individual motivation predict their grades (Level 1), and does the school's overall funding predict average school grades (Level 2)?"*

## The Example Scenario

We will use the `Demo.twolevel` dataset included in the R package `{lavaan}`, but to make it easier to follow, we will interpret the variables within an educational context.
The data contains the following information:

- **Clusters (`cluster`):** 200 different schools.
- **Observations:** 2500 students nested within these schools.
- **Outcomes (`y1`, `y2`, `y3`):** Three distinct survey items measuring "Academic Performance."
- **Within-level Predictors (`x1`, `x2`, `x3`):** Student-specific factors such as Study Hours, Sleep Hours, and Attendance.
- **Between-level Predictors (`w1`, `w2`):** School-level factors such as Teacher Experience, and School Budget.

For our model, we assume two latent variables:

1. `fw` measuring individual students aptitude (i.e. student ability)
2. `fb` measuring school quality, i.e. the shared variance in performance attributable to the school environment.

The envisaged two-level SEM can be visualized as follows:

<style>
.mermaid-js rect{
  fill: lightyellow!important;
}
</style>
```{mermaid}

graph LR

    %% --- Level 1: Within (Student) ---
    subgraph L1 [Level 1: Within-Student]
        direction LR
        x1[x1] & x2[x2] & x3[x3] --> fw((fw))
        fw --> y1_w[y1] & y2_w[y2] & y3_w[y3]
    end
    
    %% --- Separator (Invisible edge to force stacking if needed) ---
    L1 ~~~ L2

    %% --- Level 2: Between (School) ---
    subgraph L2 [Level 2: Between-School]
        direction LR
        w1[w1] & w2[w2] --> fb((fb))
        fb --> y1_b[y1] & y2_b[y2] & y3_b[y3]
    end
    
        %% --- Styling ---
    classDef latent fill:#f9f9f9,stroke:#333,stroke-width:2px,shape:circle;
    classDef observed fill:#fff,stroke:#333,stroke-width:1px,shape:rect;
    
    class fw,fb latent;
    class x1,x2,x3,w1,w2,y1_b,y2_b,y3_b,y1_w,y2_w,y3_w observed;
```


## Load the Package and Data

First, we load INLAvaan and the dataset.

```{r}
library(INLAvaan)
data("Demo.twolevel", package = "lavaan")
head(Demo.twolevel)
```

## Model Specification and Fit

In `{INLAvaan}` (following lavaan syntax), we specify the model for each level using the `level: <block>` keywords.
In our example,

- **Level 1:** We define the latent variable `fw` (Student Ability) and regress it on individual predictors (`x`).
- **Level 2:** We define the latent variable `fb` (School Quality) and regress it on school-level predictors (`w`).

```{r}
mod <- "
  level: 1
      # Measurement model (Within-student)
      fw =~ y1 + y2 + y3
      # Structural model: Individual predictors
      fw ~ x1 + x2 + x3

  level: 2
      # Measurement model (Between-school)
      fb =~ y1 + y2 + y3
      # Structural model: School-level predictors
      fb ~ w1 + w2
"
```

We use the `asem()` function (Approximate SEM) to fit the model. 
Crucially, we must specify the `cluster` argument to identify the grouping variable.

```{r}
fit <- asem(mod, data = Demo.twolevel, cluster = "cluster")
```

## Results 

The summary output provides Bayesian estimates (posterior means, standard deviations, and credible intervals) for *both levels*.

```{r}
summary(fit)
```

Notice that, the mean structure is automatically included at both levels, so intercepts for all observed variables are estimated by default.
This is required because the 'between' component specifically models the variation of the cluster means; without estimating these means (intercepts), it is impossible to decompose the variance into within and between levels.
Looking at the output above, we can draw substantive conclusions based on our educational scenario:

```{r}
#| include: false
est <- coef(fit)
fmt <- function(x) sprintf("%.3f", x)
```


- **Level 1 [within] Regressions**
  
  The path `fw ~ x1` is `r fmt(est["fw~x1"])`. This suggests that for every unit increase in `x1` (e.g., Study Hours), the student's individual ability (`fw`) increases significantly.

- **Level 2 [cluster] Regressions**

  The path `fb ~ w1` is `r fmt(est["fb~w1.l2"])` This suggests a positive relationship between school-level factors (like Teacher Experience) and the overall School Quality (`fb`), though the standard deviation is wider here due to the smaller sample size at Level 2 ($n=200$ schools vs $n=2500$ students).

- **Latent Variables:**
  
  The loadings for `y1`, `y2`, and `y3` on both `fw` and `fb` are significant (0 not included in credible interval) and thus confirm that these survey items effectively measure both individual ability and school-level quality.




## Random Slopes

The model above lets every school have its own average performance, but it forces the *effect* of study hours to be identical in all 200 schools.
Often that is exactly the question of interest: does an extra hour of study buy more in some schools than in others?
A **random slope** answers it by promoting a Level 1 regression coefficient to a Level 2 latent variable, with a mean, a variance, and cluster-level predictors of its own.

In lavaan syntax a slope is made random by wrapping it in the `rv()` modifier, which gives it a name.
That name is then an ordinary Level 2 latent variable, so it can be regressed on school-level predictors -- here on teacher experience `w1`, asking whether the study-hours effect is larger in schools with more experienced teachers.

```{r}
mod_rs <- "
  level: 1
      # Measurement model (Within-student)
      fw =~ y1 + y2 + y3
      # The effect of x1 is now a latent variable called s1
      fw ~ rv('s1')*x1

  level: 2
      # Measurement model (Between-school)
      fb =~ y1 + y2 + y3
      fb ~ w1
      # s1 behaves like any other Level 2 latent variable
      s1 ~ w1
"
```

The mean and variance of `s1` are added automatically, so only the cross-level regression has to be written out.
The fit is requested exactly as before.

```{r}
fit_rs <- asem(mod_rs, data = Demo.twolevel, cluster = "cluster")
```

```{r}
summary(fit_rs)
```

Two things in that output are specific to random slopes.
At Level 1 the carrier row for `x1` is tagged `(s1)` and fixed at zero: the slope is no longer a Level 1 parameter, so there is nothing to report there.
At Level 2 the slope appears in its own right -- its mean under *Intercepts* (`.s1`, the average study-hours effect across schools), its variance under *Variances* (`.s1`, how much that effect varies between schools), and its regression on `w1` under *Regressions*.

### Fit Measures

```{r}
fitmeasures(fit_rs)
```

The posterior predictive p-value is missing.
For an ordinary two-level model it scores replicate data sets against the saturated model, as blavaan does (by default with one Fisher-scoring step towards the saturated fit of each replicate; `ppp_method = "em"` gives blavaan's full EM fit), but a random-slope model has no saturated model: the covariance of the outcomes depends on the values the covariates happen to take, so an unrestricted model would need its own covariance matrix for every school.
INLAvaan therefore drops `"ppp"` from the default `test` (the message printed above the fit).

The Bayesian fit indices are present, on a different footing.
Their chi-square is taken against the most general model with the same random effects -- every outcome with its own random intercept and its own random slope on `x1`, all freely correlated and all regressed on `w1`, and an unrestricted within-school residual covariance -- in which the fitted model is nested.
That reference is INLAvaan's own construction rather than an established standard, so the indices say how close the model comes to it, and it needs enough schools to estimate its parameters.
It cannot be built for a model with a between-only outcome, because lavaan's random-slope kernel allows such an outcome only as an indicator of a latent variable.
The model-comparison machinery below is what answers the substantive question.

### Is There a Random Slope at All?

The Bayesian answer is a model comparison.
The natural comparator fixes the slope variance at zero **and** drops the cross-level regression on the slope: what is left is a slope that is one constant shared by every school, which is exactly the fixed-slope model, with the same log-likelihood and the same number of parameters.
Keeping `s1 ~ w1` while fixing the variance gives a useful model in between -- a slope that varies from school to school, but deterministically, as a cross-level interaction with teacher experience.

```{r}
mod_cli <- "
  level: 1
      fw =~ y1 + y2 + y3
      fw ~ rv('s1')*x1

  level: 2
      fb =~ y1 + y2 + y3
      fb ~ w1
      s1 ~ w1
      # A school-varying slope, but with no variance of its own
      s1 ~~ 0*s1
"

mod_fx <- "
  level: 1
      fw =~ y1 + y2 + y3
      fw ~ rv('s1')*x1

  level: 2
      fb =~ y1 + y2 + y3
      fb ~ w1
      # One constant slope shared by every school
      s1 ~~ 0*s1
"
fit_cli <- asem(mod_cli, data = Demo.twolevel, cluster = "cluster")
fit_fx <- asem(mod_fx, data = Demo.twolevel, cluster = "cluster")
```

```{r}
cmp <- compare(fit_rs, fit_cli, fit_fx)
cmp
```

```{r}
#| include: false
bf <- stats::setNames(cmp$logBF, cmp$Model)
```

Because a random-slope likelihood is the density of the outcomes *given* the covariates, every fit in a table like this one must score the same variables: the outcomes, plus any between-level variable the model explains rather than regresses on.
`compare()` checks that, and aborts otherwise.
The covariates themselves may differ, as for any fit with `fixed.x = TRUE`: a model without `w1` says that the outcomes do not depend on it, so dropping a covariate is a fair comparison too.
The plain fixed-slope model written out directly (`fw ~ x1` at Level 1) is an equally valid member of the table, and it fits a good deal faster.

On `Demo.twolevel` the maximum-likelihood estimate of the slope variance is slightly negative, so there is no slope variation in these data to find, and the models without it are duly preferred: a log Bayes factor of `r fmt(bf["fit_cli"])` for the cross-level-interaction model and `r fmt(bf["fit_rs"])` for the full random-slope model, both against the fixed-slope fit, with the DIC ordering the three the same way.

### Cross-Validation

`loo()` and `waic()` work on random-slope fits too.

```{r}
loo(fit_rs)
```

The units are always clusters (leave-one-cluster-out), and the score is always the *conditional* flavour: the likelihood is the density of the outcomes given the covariates, so what is cross-validated is the prediction of a new school's outcomes from its covariates.
Leave-one-observation-out is not available here, because deleting a single row would need a school's sufficient statistics downdated by one row and the random-slope likelihood is not built from such statistics.

```{r}
compare(fit_rs, fit_fx, loo = TRUE)
```

The paired standard error is the right uncertainty for a nested comparison like this one, and it confirms the reading above: the fixed-slope model predicts a held-out school slightly better, by about one elpd unit.

### School-Level Slopes

Even when the variance is small, posterior draws of the school-specific slopes are available, alongside the other Level 2 latent variables.

```{r}
predict(fit_rs, type = "lv", level = 2, summary = TRUE)
```

### Implied Moments and Standardised Estimates

A random-slope model has no single within-cluster covariance matrix, and lavaan's own implied moments simply leave the slope out.
INLAvaan's `fitted()` instead averages the implied moments over the covariates, holding them at their sample moments as `fixed.x = TRUE` does.
Each slope then contributes its mean, as a fixed slope would, and its variance times the variance of its covariate, which is the random-slope component of the within-cluster variance in the framework of @rights2019quantifying.
With the slope variance at zero and no cross-level regression on the slope, these are exactly the moments of the fixed-slope model.

```{r}
fitted(fit_rs)$within
```

`residuals()` compares these with the sample moments as usual.
With `per_cluster = TRUE`, both functions work school by school instead: the expected mean and within-school covariance of each school at its own covariate values, against that school's own sample moments.

```{r}
residuals(fit_rs, per_cluster = TRUE)[["1"]]
```

Standardised estimates are scaled by the averaged variances.
The Level 1 carrier row `x1 (s1)` then shows the standardised mean slope, and the Level 2 rows of `s1` are on the same standardised-slope scale: the intercept of `s1` is a standardised slope, and its residual variance is the share of the within-school variance of `fw` that the slope's variation adds beyond what `w1` explains.
Without a cross-level regression on the slope, its square root is the standard deviation of the standardised slopes across schools.

```{r}
std <- standardisedsolution(fit_rs)
std[std$lhs %in% c("fw", "s1"), ]
```

### What Is Not Available

INLAvaan prefers an error that explains itself to a plausible wrong number.
The posterior predictive p-value is dropped, as described above, and the following raise an error on a random-slope fit:

- residuals scaled by standard errors (`type = "normalized"` or `"standardized"`);
- `predict(type = "ymis")`;
- `loo(type = "loso")`.

Everything else works.
`simulate()` draws replicate data sets in which every school keeps its own covariates and size, `sampling()` gives latent, observed and implied draws, `predict(type = "yhat")` gives each student's predicted outcomes from the school's own random effects, and `fitted(type = "casewise")` gives the outcomes' means given the covariates alone.

A random-slope model with observed exogenous covariates also requires `fixed.x = TRUE` (lavaan's default): the likelihood conditions on those covariates, so their means and (co)variances are unidentified and would simply be reported back as their priors.
A model whose covariates are all latent or modelled has nothing to hold fixed, and lavaan reports `fixed.x = FALSE` for it of its own accord; such a fit is accepted as it stands.

### Latent and Split Covariates

Everything above uses a covariate that is observed and purely within-cluster, for which the random slope integrates out in closed form.
When the covariate is latent, or split across both levels (the same variable entering at Level 1 and at Level 2), the integral has to be done by Gauss-Hermite quadrature instead [@rockwood2020maximum].
INLAvaan takes that route automatically and warns when it does.
The argument `integration.ngh` sets the number of nodes per dimension; because it is an accuracy setting as much as a cost setting, every fit that is to appear in the same `compare()` table must use the same value, which `compare()` also checks.
A split covariate is conditioned on just as a purely within-cluster one is, so the plain fixed-slope model with the same covariates is the comparator to use; fixing the slope variance at zero is not possible on this route.
The averaged moments, the standardised estimates and the `sampling()` draws are available on this route too.
The data generator of `simulate()`, `per_cluster = TRUE`, the casewise and `yhat` values and the Bayesian fit indices are not, because each school's outcomes are then a mixture over the quadrature nodes.

## References
