---
title: 'Multigroup Analysis'
description: >
  Measurement invariance testing with multigroup SEM: the configural,
  metric, and scalar ladder, compared honestly with leave-one-out
  cross-validation.
vignette: >
  %\VignetteIndexEntry{Multigroup Analysis}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
bibliography: ../refs.bib
toc: true
---

```{r}
#| label: setup
#| include: false
library(INLAvaan)
library(lavaan)
```

## The measurement invariance ladder

Multigroup SEM fits the same model structure to several groups at once,
with each group keeping its own parameters unless told otherwise.
The classic application is **measurement invariance** [@meredith1993measurement]:
does an instrument measure the same construct, on the same scale, across
groups?
The question is asked as a ladder of increasingly constrained models:

- **Configural**: same factor structure in every group, all parameters free.
- **Metric** (weak): loadings constrained equal across groups -- the latent
  variables are on a common scale, so latent *covariances* can be compared.
- **Scalar** (strong): loadings and intercepts equal -- observed scores map
  to latent scores the same way, so latent *means* can be compared.

Each rung is fitted by adding to `group.equal`.
We use the Holzinger--Swineford data, where 301 children from two schools
(Pasteur and Grant-White) took nine cognitive tests:

```{r}
#| label: fit-ladder
#| message: false
HS.model <- "
  visual  =~ x1 + x2 + x3
  textual =~ x4 + x5 + x6
  speed   =~ x7 + x8 + x9
"
data("HolzingerSwineford1939", package = "lavaan")

fit_configural <- acfa(HS.model, HolzingerSwineford1939, group = "school",
                       meanstructure = TRUE, verbose = FALSE)
fit_metric <- acfa(HS.model, HolzingerSwineford1939, group = "school",
                   group.equal = "loadings",
                   meanstructure = TRUE, verbose = FALSE)
fit_scalar <- acfa(HS.model, HolzingerSwineford1939, group = "school",
                   group.equal = c("loadings", "intercepts"),
                   meanstructure = TRUE, verbose = FALSE)
```

Cross-group equality constraints are handled exactly (not by penalty):
the constrained loadings become single free parameters shared by both
groups.
A `summary()` of any fit reports the parameters per group.

## Comparing the rungs

`compare()` reports marginal log-likelihoods, Bayes factors, and DIC:

```{r}
#| label: compare-bf
compare(fit_configural, fit_metric, fit_scalar)
```

These are within-flavour comparisons (all three fits model the means with
proper priors), so the Bayes factors are meaningful here.
But marginal likelihoods answer a prior-sensitive question -- "which model
gave these data the highest prior predictive density?" -- and for
invariance testing we usually want the predictive one: *which rung of the
ladder predicts new children best?*

`compare(..., loo = TRUE)` answers that with leave-one-out
cross-validation computed from each single fit -- no refitting, no
sampling (see the [LOO article](loo.html) for the machinery).
Groups are independent, so every child is scored against their own
school's implied moments, and the expected log predictive density (ELPD)
sums over both schools.
Differences come with *paired* standard errors, since the same children
are scored under every model:

```{r}
#| label: compare-loo
compare(fit_configural, fit_metric, fit_scalar, loo = TRUE)
```

The textbook Holzinger--Swineford story appears: metric invariance is
indistinguishable from configural (the paired difference is well within
one standard error, so the equal-loadings model -- simpler, with the same
predictive reach -- is preferred), while scalar invariance costs over
two standard errors of ELPD: the intercepts genuinely differ between
schools.
Note the instructive disagreement with the Bayes factors, which reward
prior-predictive parsimony and pick the scalar model; on predictive
grounds the ladder stops at metric.

## Step zero: is there any group difference at all?

The ladder presumes grouping matters.
That too is a model comparison: the pooled (single-group) model against
the configural one.
Units are identified by their case number -- the row of the dataset --
not by their position in a group, so `compare()` pairs the pooled and
grouped fits child by child even though the two fits order the data
differently:

```{r}
#| label: compare-pooled
#| message: false
fit_pooled <- acfa(HS.model, HolzingerSwineford1939,
                   meanstructure = TRUE, verbose = FALSE)
compare(fit_pooled, fit_configural, loo = TRUE)
```

Allowing the two schools their own parameters clearly improves
prediction, so the multigroup analysis is warranted.

## Pointwise diagnostics by group

`loo()` on a multigroup fit returns the usual per-unit table with a
`group` column, which makes group-level diagnostics one `tapply` away --
for instance, how each school contributes to the scalar rung's loss:

```{r}
#| label: loo-by-group
loo_metric <- loo(fit_metric)
loo_scalar <- loo(fit_scalar)
loo_metric

head(loo_metric$per_unit[, 1:6], 3)

# Where does scalar invariance lose predictive density?
d <- loo_scalar$per_unit$log_cpo_2 - loo_metric$per_unit$log_cpo_2
tapply(d, loo_scalar$per_unit$group, sum)
```

Both schools pay for the intercept constraints, confirming the misfit is
not driven by a handful of cases in one school.

`waic()` works the same way for multigroup fits, with the same `group`
column in its pointwise table.

## Means or no means

Measurement invariance needs modelled means from the scalar rung up, but
the lower rungs can also be fitted without a mean structure
(`meanstructure = FALSE`, the marginalised-means treatment).
Marginal likelihoods and Bayes factors are not comparable across the two
mean treatments, while LOO remains valid across that boundary -- each
child's leave-one-out conditional is a proper density under either
treatment, with the exchangeability transformation applied within each
group.
See the [mean structures article](meanstructure.html) for the full story.

## References
