---
title: 'Sampling from the Generative Model'
description: >
  Draw model parameters, latent variables, and observed data from the generative SEM—useful for posterior and prior predictive checks.
vignette: >
  %\VignetteIndexEntry{Sampling from the Generative Model}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk: 
    collapse: true
    comment: '#>'
    fig.width: 7
    fig.height: 4.5
bibliography: ../refs.bib
toc: true
---

```{r}
#| label: setup
#| include: false
devtools::load_all()
library(ggplot2)
set.seed(42)
```

## Purpose

After fitting a Bayesian SEM with INLAvaan, we often want to **draw samples from the model** itself—either using posterior or prior parameter values.
The `sampling()` function does exactly this: it propagates parameter draws through the full generative chain

$$
\underbrace{\boldsymbol\theta}_{\text{parameters}}
\;\longrightarrow\;
\underbrace{\boldsymbol\eta}_{\text{latent variables}}
\;\longrightarrow\;
\underbrace{\mathbf{y}^*}_{\text{observed variables}}
$$

producing samples that are **not tied to any individual observation**.
This is distinct from `predict()`, which returns individual-specific factor scores $\boldsymbol\eta \mid \mathbf{y}, \boldsymbol\theta$.

Typical use cases include **posterior predictive checks** (PPCs) and **prior predictive checks**.


## The generative model

Let $\boldsymbol\theta^{(s)}$ ($s = 1, \dots, S$) denote one parameter draw.
From this draw the SEM matrices $\boldsymbol\Lambda$, $\boldsymbol\Psi$, $\mathbf{B}$, $\boldsymbol\alpha$, $\boldsymbol\nu$, and $\boldsymbol\Theta$ are constructed.
The generative chain is:

**1. Latent variables.**
$$
\boldsymbol\eta^{(s)}
\sim
\mathcal{N}\!\bigl(
  (\mathbf{I} - \mathbf{B})^{-1}\boldsymbol\alpha,\;
  \boldsymbol\Phi
\bigr),
\qquad
\boldsymbol\Phi
= (\mathbf{I} - \mathbf{B})^{-1}\boldsymbol\Psi\,
  [(\mathbf{I} - \mathbf{B})^{-1}]'.
$$

**2. Observed variables.**
$$
\mathbf{y}^{*(s)}
\sim
\mathcal{N}\!\bigl(
  \boldsymbol\Lambda\,\boldsymbol\eta^{(s)} + \boldsymbol\nu,\;
  \boldsymbol\Theta
\bigr).
$$

When `prior = FALSE` (default), $\boldsymbol\theta^{(s)}$ comes from the posterior; when `prior = TRUE`, each parameter is drawn independently from its prior.


## Quick start

```{r}
#| label: fit-model
dat <- lavaan::HolzingerSwineford1939
mod <- "
  visual  =~ x1 + x2 + x3
  textual =~ x4 + x5 + x6
  speed   =~ x7 + x8 + x9
"
fit <- acfa(mod, dat, verbose = FALSE)
```

### Parameter samples

The default type `"lavaan"` returns an $S \times p$ matrix of lavaan-side (constrained) parameter draws:

```{r}
#| label: param-samples
theta_post <- sampling(fit, type = "lavaan", nsamp = 2000)
dim(theta_post)
head(theta_post[, 1:4])
```

### Latent and observed samples

```{r}
#| label: latent-observed
eta  <- sampling(fit, type = "latent",   nsamp = 1000)
ystar <- sampling(fit, type = "observed", nsamp = 1000)
dim(eta)    # 1000 x 3 (one per latent factor)
dim(ystar)  # 1000 x 9 (one per observed indicator)
```

### Everything at once

```{r}
#| label: all
all_samps <- sampling(fit, type = "all", nsamp = 1000)
names(all_samps)
```


## Comparing samples to the fitted marginals

INLAvaan approximates each marginal posterior with a skew-normal density.
We can overlay the `sampling()` histogram (drawn via the copula method) on top of the fitted density curve stored in `pdf_data` to verify that they agree.

```{r}
#| label: fig-copula-vs-pdf
#| fig-cap: "Copula samples (histogram) versus the fitted skew-normal marginal (red curve) for all free parameters."
# 1. Draw copula samples
samp_cop <- sampling(fit, type = "lavaan", nsamp = 5000, samp_copula = TRUE)

# 2. Retrieve the fitted skew-normal densities
int <- fit@external$inlavaan_internal
pdf_data <- int$pdf_data

# 3. Build a long data frame for ggplot
par_names <- colnames(samp_cop)
hist_df <- data.frame(
  param = rep(par_names, each = nrow(samp_cop)),
  value = as.vector(samp_cop)
)
hist_df$param <- factor(hist_df$param, levels = par_names)

curve_df <- do.call(rbind, Map(function(nm, df) {
  data.frame(param = nm, x = df$x, y = df$y)
}, par_names, pdf_data[par_names]))
curve_df$param <- factor(curve_df$param, levels = par_names)

# 4. Plot
ggplot(hist_df, aes(x = value)) +
  geom_histogram(aes(y = after_stat(density)), bins = 50,
                 fill = "steelblue", alpha = 0.5) +
  geom_line(data = curve_df, aes(x = x, y = y),
            colour = "firebrick", linewidth = 0.8) +
  facet_wrap(~param, scales = "free") +
  labs(x = "Parameter value", y = "Density",
       title = "Copula samples vs. fitted skew-normal posterior") +
  theme_minimal(base_size = 11)
```


## Copula vs. Gaussian sampling

By default, `sampling()` uses the copula method, which respects the skew-normal marginals.
Setting `samp_copula = FALSE` uses the multivariate Gaussian (Laplace) approximation instead.
The difference is most visible for parameters with asymmetric posteriors, such as variance components.

```{r}
#| label: fig-copula-vs-gaussian
#| fig-cap: "Comparison of marginal histograms from copula sampling (blue) and Gaussian sampling (orange) for all parameters."
samp_gauss <- sampling(fit, type = "lavaan", nsamp = 5000, samp_copula = FALSE)

cop_df <- data.frame(
  param = rep(par_names, each = nrow(samp_cop)),
  value = as.vector(samp_cop),
  method = "Copula"
)
gauss_df <- data.frame(
  param = rep(par_names, each = nrow(samp_gauss)),
  value = as.vector(samp_gauss),
  method = "Gaussian"
)
both_df <- rbind(cop_df, gauss_df)
both_df$param <- factor(both_df$param, levels = par_names)
both_df$method <- factor(both_df$method, levels = c("Copula", "Gaussian"))

ggplot(both_df, aes(x = value, fill = method)) +
  geom_histogram(aes(y = after_stat(density)), bins = 50,
                 alpha = 0.45, position = "identity") +
  facet_wrap(~param, scales = "free") +
  scale_fill_manual(values = c(Copula = "steelblue", Gaussian = "darkorange")) +
  labs(x = "Parameter value", y = "Density", fill = "Method",
       title = "Copula vs. Gaussian posterior sampling") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top")
```

For symmetric posteriors (e.g., factor loadings), the two methods are nearly identical.
For positively-skewed posteriors (e.g., residual variances), the copula method captures the asymmetry while the Gaussian approximation is symmetric by construction.


## Prior predictive sampling

Setting `prior = TRUE` draws parameters from the prior distribution (independently, ignoring the data) and propagates them through the generative model.
This is useful for checking whether the priors imply sensible ranges for the observed data.

```{r}
#| label: prior-predictive
y_prior <- sampling(fit, type = "observed", nsamp = 2000, prior = TRUE)
y_post  <- sampling(fit, type = "observed", nsamp = 2000, prior = FALSE)
```

```{r}
#| label: fig-prior-post-predictive
#| fig-cap: "Model-implied distribution of x1 under the prior (orange) and posterior (blue)."
df_pp <- data.frame(
  x1 = c(y_prior[, "x1"], y_post[, "x1"]),
  source = rep(c("Prior", "Posterior"), each = 2000)
)

ggplot(df_pp, aes(x = x1, fill = source)) +
  geom_histogram(aes(y = after_stat(density)), bins = 50,
                 alpha = 0.5, position = "identity") +
  scale_fill_manual(values = c(Prior = "darkorange", Posterior = "steelblue")) +
  labs(x = "x1", y = "Density", fill = NULL,
       title = "Prior vs. posterior predictive distribution of x1") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top")
```
